---
title: "MS_Thesis_power_analysis"
author: "Nick Foreman"
date: "3/5/2021"
output: html_document
---

First, I'll load the data from a previous study by Peake et al. (2014), which compared the effects of moderate and high intensity exercise on metabolism after exercise. The dataset is a summary of the means and SDs of the different metabolites in this study. They used a slightly different analysis method than I'll be using, but it's comparable enough for the power analysis to work. 

```{r}
library(pwr2); library(tidyverse)

data <- read_csv(file = "peake_sum_data.csv")

#Factor A is Protocol (MOD vs. HIIT)
#Factor B is Timepoint (POST vs. 1 H)

metabs <- colnames(data)[4:51]
all_metabs <- list()

```

For each metabolite in the Peake et al. data, this calculates a pooled SD and divides the difference in means by the pooled SD to find an effect size. This effect size is what I used to find sample size. The pwr2 package documentation says they use methods from a Springer text by Dean & Voss (1999). The number of samples (n) required for each group in the 2x2 ANOVA is reported as part of the output and stored in a list with the metabolite name. 

```{r}
for (i in 1:length(metabs)){
  #Prepare data for factor A
  f_A <- data %>% 
    drop_na() %>% #get rid of rows with calculations 
    filter(Timepoint == "POST" | Timepoint == "1 H") %>%
    select(1:3, contains(metabs[i])) %>%
    filter(Timepoint=="POST") %>% 
    pull(metabs[i])
  
  #Factor A effect size
  sd_pool_A <- sqrt(((f_A[2])^2 + (f_A[4])^2)/2)
  mean_diff_A <- (f_A[1] - f_A[3])
  effect_size_A <- mean_diff_A/sd_pool_A
  
  f_B <- data %>% 
    drop_na() %>% #get rid of rows with calculations 
    filter(Timepoint == "POST" | Timepoint == "1 H") %>%
    select(1:3, contains(metabs[i])) %>%
    #filter(Timepoint=="POST") %>% 
    filter(Protocol == "MOD") %>%
    pull(metabs[i])
  
  #Factor B effect size
  sd_pool_B <- sqrt(((f_B[3])^2 + (f_B[4])^2)/2)
  mean_diff_B <- abs(f_B[1]-f_B[2])
  effect_size_B <- mean_diff_B/sd_pool_B
  
  results <- ss.2way(a=2, b=2, alpha=0.05, beta=0.2, #80% power
                     delta.A = NULL, delta.B = NULL,
                     sigma.A = NULL, sigma.B = NULL,
                     f.A = effect_size_A, f.B = effect_size_B,
                     B=100) 
  #B is the number of iterations, 100 is the default, doesn't seem to change result
  
  all_metabs[[i]] <- metabs[i]
  names(all_metabs[[i]]) <- metabs[i]
  all_metabs[[i]][[1]] <- as.numeric(results$n)
  
}

```

This returns information about the number N in each group for a two-way ANOVA. I'm not certain that a two-way ANOVA is best for me, but I figured it was a good place to start. As you can see from this rough histogram, there's some variety in the number of samples per group, from about 5 to about 100. I'm not quite sure how to handle this, as it's somewhat unique to metabolomics and not something I would find in other lower-dimensionality datasets I've worked with.

```{r}
hist(as.numeric(unlist(all_metabs)),breaks = 15)

```
I currently have funding for 10 participants, and I hope to have funding for 20. 
This outputs a count of the number of metabolites I have sufficient power for (n=10), might be able to get funding for (n=15), would really have to work to get funding for (n=20), and have no way of getting funding for/don't have enough time in a thesis anyways (n>20) 

```{r}
n_sufficient_pwr <- which(all_metabs <= 10)
#all_metabs[n_sufficient_pwr]
length(all_metabs[n_sufficient_pwr])

n_reasonable_pwr <- which(all_metabs <= 15)
#all_metabs[n_sufficient_pwr]
length(all_metabs[n_reasonable_pwr])

n_maybe_pwr <- which(all_metabs <= 20)
#all_metabs[n_sufficient_pwr]
length(all_metabs[n_maybe_pwr])

n_insufficient_pwr <- which(all_metabs > 20)
#all_metabs[n_insufficient_pwr]
length(all_metabs[n_insufficient_pwr])
```